Viscous Erosion in Porous Media

### 2018-02-22 docs/methods/erosion_methods.tex
NM

\todo[inline]{Note: I took out 'in the Stokesian limit' because the vorticity matches the shear stress on the boundary no matter what the Reynolds number is.}

@@

With the shear stress computed, we are now ready to evolve the interface
of each body using equations~\eqref{eqn:Levo1}
and~\eqref{eqn:thetaevo1}. As described in Section~\ref{sec:thetaL}, we
use the {\thL} framework, with $\theta$ being the tangent angle and $L$
the total arc length of a given boundary. Each boundary is
parameterized by normalized arc length $\alpha = 2 \pi s / L$.

@@

We will evolve $\theta(\alpha)$ spectrally and so introduce the Fourier series
\begin{align}
\label{thetaSeries}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha}, \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha}.
\end{align}
The careful reader will notice that in \eqref{thetaSeries}, we have decomposed $\theta(\alpha)$ into a linear component and a periodic component.
In spectral space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k.
\end{align*}
Each ODE is first-order and linear (though variable-coefficient due to
$\elfun$) and thus can be solved by using the integrating factor
\begin{align}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right).
\end{align}
We are therefore led to the coupled system of ODEs
\begin{align}
\label{eqn:Levo3}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo3}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k \, ,
\qquad k = -N_\iin/2, \ldots, N_\iin/2 -1
\end{align}
We solve system~\eqref{eq:mu}--\eqref{eqn:thetaevo3} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we choose the {\em midpoint} RK2 method so as to maximize the smoothing effect of the curvature term.

At the first stage of RK2 (the half step) we discretize \eqref{eq:mu}--\eqref{eqn:thetaevo3} as
\begin{align}
\label{eq:mu12d}
& \mu_k^{(n+1/2)} = \mu_k^{(n)} \exp \left( \eps k^2 \frac{\Dt}{2} \zeta^{(n)} \right) \\
\label{eq:L12d}
& \frac{L^{(n+1/2)} - L^{(n)}}{\Dt/2} = \Mterm^{(n)} \\
\label{eq:th12d}
& \frac{ \mu_k^{(n+1/2)} \thhat_k^{(n+1/2)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt/2} 
= \mu_k^{(n)} \hat{\NLterm}_k^{(n)}
\end{align}
where the superscript indicates the time step. At the next stage of RK2 it will be necessary to evaluate derivates at the half step $n+1/2$, thus requiring a call to the Stokes solver for which the shape of the bodies must be known through $\theta^{(n+1/2)}$. We therefore use \eqref{eq:mu12d} to solve \eqref{eq:th12d} for $\theta^{(n+1/2)}$, giving
\begin{align}
\label{eq:th12}
\thhat_k^{(n+1/2)} = \left( \thhat_k^{(n)} + \frac{1}{2} \Dt \, \hat{\NLterm}_k^{(n)} \right)
\exp \left( -\epsilon k^2 \frac{\Dt}{2} \elfun^{(n)}\right)
\end{align}
Since $\zeta >0$, the exponential term always acts to smooth the geometry and thus mitigate numerical instabilities. With the geometry at step $n+1/2$ known, a call to the Stokes solver allows one to compute the terms $\Mterm^{(n+1/2)}$, $\widehat{\NLterm}_k^{(n+1/2)}$, and $\zeta^{(n+1/2)}$.

At the second stage of RK2, we discretize \eqref{eq:mu}--\eqref{eqn:thetaevo3} as 
\begin{align}
\label{eq:mu1d}
& \mu_k^{(n+1)} = \mu_k^{(n)} \exp \left( \eps k^2 {\Dt} \, \zeta^{(n+1/2)} \right) \\
\label{eq:L1d}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)}, \\
\label{eq:th1d}
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt} =
\mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}.
\end{align}
Once again, we must eventually compute the geometry at step $n+1$, and so we use \eqref{eq:mu1d} to solve \eqref{eq:th1d} for $\theta^{(n+1)}$, giving
\begin{align}
\thhat_k^{(n+1)} = & \thhat_k^{(n)} \exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) + \Dt \, \hat{\NLterm}_k^{(n+1/2)} \exp \left( - \frac{1}{2} \eps k^2 \Dt \left( 2 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right).
\end{align}
Again, both exponential terms act to reduce high-frequency oscillations in shape. We note that the above approach would generalize to higher-order Runge-Kutta methods, such as RK4.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Vorticity}
\label{sec:vorticity}
We start by computing the vorticity $w(\xx) = v_x - u_y$ for $\xx \in
\Omega$.  As we did for the shear stress, we compute the vorticity due
to the double-layer potential, Stokeslets, and rotlets individually.
The vorticity of the double-layer potential~\eqref{eqn:velocityDLP} at
$\xx \in \Omega$ is
\begin{align}
  w^{\DD}(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} 
    \frac{(\rr \cdot \nn^\perp) + (\rr \cdot \nn)}{\rho^4}
    (\rr \cdot \eeta) ds_{\yy}.
  \label{eqn:vorticityDLP}
\end{align}
The vorticity due to the Stokeslet is
\begin{align*}
  w^S(\xx) = -\frac{1}{\pi} \sum_{\ell=1}^{M} 
    \frac{\rr \cdot \llambda_\ell^\perp}{\rho^2},
\end{align*}
and the vorticity of a rotlet is zero.  Therefore, the vorticity at $\xx
\in \Omega$ is
\begin{align*}
  w(\xx) = w^\DD(\xx) + w^S(\xx).
\end{align*}
To compute~\eqref{eqn:vorticityDLP} numerically, we apply the trapezoid
rule with an upsampling scheme described in Section~\ref{sec:NSI}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Drag}
\label{sec:drag}
\todo[inline]{check all signs in this section.  Fortran code assumes a
normal pointing into the eroding body and a tangent vector that
corresponds to a couterclockwise parameterization}
With the pressure and shear stress, the drag on a single body with
boundary $\gamma_\ell$ is
\begin{align}
  F_D = \int_{\gamma_\ell} (-p \nn + \tau \ss)\,ds.
  \label{eqn:drag}
\end{align}

@@

Fig.~\ref{01bodseq} shows the erosion of a single body, numerically
simulated with $N_\iin = 1024$ discretization points and shown at three
evenly spaced times. Here, color represents the vorticity of the
surrounding flow, which, when evaluated on the boundary, indicates the
shear stress and thus the local erosion rate. As seen in the left-most
snapshot, the body is initially circular and has the strongest vorticity
above and below it. The flow is for-aft symmetric as a result of the
time-reversibility of the Stokes equations, with vorticity vanishing at
the body's front and rear stagnation points. At the next instance, the
body has lost material due to erosion, but remains for-aft symmetric due
to the flow symmetry. The greatest material loss has occurred at the
body's top and bottom, where shear is strongest. Interestingly, the body
has developed well-defined corners at its front and rear stagnation
points. This geometric feature is a consequence of the erosion law,
which depends on the {\em absolute} shear stress: wherever the stress
vanishes, $\atau$ exhibits a corner, which manifests as a corner in the
geometry. As shown in the third snapshot, this corner persists as the
bodies becomes even more slender while wasting away. In this simulation,
the body completely vanishes at a final time of $t_f = 1.79E-2$.

To better understand the shape evolution, Fig.~\ref{shrink_intface}(a)
shows several superimposed interfaces at evenly spaced times. The
interfaces are color-coded by normalized time $t/t_f$. The body seems to
converge to a shape that resembles an American football---a fairly
slender, for-aft symmetric shape with corners at the front and back.
Fig.~\ref{shrink_intface}(b) shows how the absolute shear stress,
$\atau$, is distributed along each of these interfaces. The normalized
arc length $s/L$ runs counterclockwise around the body, with $s/L = 0$
corresponding to the rear stagnation point and $s/L = 0.5$ the front
stagnation point. At early times (yellow to orange), the stress is
distributed non-uniformly, with highest values at the top and bottom of
the body. As time proceeds, though, the stress becomes more evenly
distributed along the body's surface (red). As the body nears the final
stages of erosion (dark red), the stress actually becomes slightly
higher near the stagnation points. A close look reveals that {\em at}
the stagnation points, the stress always vanishes, though in a region
that becomes increasingly narrow.

@@

Interestingly, a brief look at Fig.~\ref{shrink_intface}(a) reveals that
the spacing between successive interfaces becomes larger as the body
shrinks, suggesting that the overall shear stress is growing. Of course,
such a relationship is consistent with the well-known 3D Stokes drag
law, $\tau \sim \mu \umax / L$, since stress is inversely proportional
to body scale $L$. This law must be modified in two dimensions, though.
The modification (made non-trivial by subtleties of the Stokes paradox)
can be correctly deduced by noting that, as the body vanishes, its drag
must be consistent with slender-body theory. For a slender body in flow
perpendicular to its long-axis $W \gg L$, the drag is proportional to
$\mu \umax W/ \log(W/L)$. This relationship implies that, for a 2D
cross-section perpendicular to the long axis, the stress (per unit $W$)
can be estimated by
\begin{align}
  \label{eqn:stresslaw}
  \tau \sim \frac{\mu \umax}{L \log(W/L)}.
\end{align}
In interpreting our 2D simulations, we will treat $W$ as fixed with the
requirement $W \gg L$.

Now, the body's rate of area reduction is given by integrating the
absolute shear stress around the boundary
\begin{align*}
  \dot{A} = \int_{\gamma} \atau \, ds \, ,
\end{align*}
giving the exact relationship
\begin{align*}
  \dot{A} = \mean{\atau} L \, .
\end{align*}
Thus, the area-loss rate is given by the product of the mean shear
$\mean{\atau}$ and the total perimeter $L$. Inserting the scaling
law~\eqref{eqn:stresslaw} gives
\begin{align*}
  \dot{A} \sim \frac{\mu \umax}{\log(W/L)} \, .
\end{align*}
Noting that $A \sim L^2$ allows one to solve the ODE to determine how the body area changes with time
\begin{align*}
  A (1 - \log A) = \areaconst (t_f - t)
\end{align*}
where $c_A$ is a constant.
This is an {\em implicit} relationship for how the area scales down in time during erosion.
In Fig.~6a, we show the body area as a function of time measured in the simulation
against the above scaling law with $\areaconst = 0.42$.

>>>>

@@

With the shear stress computed, we are now ready to evolve the boundary of each body using equations~\eqref{eqn:Levo1} and~\eqref{eqn:thetaevo1}.  As described in Section~\ref{sec:thetaL}, we use the {\thL} framework, with $\theta$ being the tangent angle and $L$ the total arc length of a given boundary. Each boundary is parameterized by normalized arc length $\alpha = 2 \pi s / L$.

@@

\todo[inline]{Need to remind reader that the diffusive term comes from smoothing and that this is the stiff term that could potential cause numerical problems.}

We will evolve $\theta(\alpha)$ spectrally and so introduce the Fourier series
\begin{align}
\label{thetaSeries}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha}, \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha}.
\end{align}
The careful reader will notice that in \eqref{thetaSeries}, we have decomposed $\theta(\alpha)$ into a linear component and a periodic component.
In spectral space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k.
\end{align*}
These ODEs can be solved by using the integrating factor
\begin{align}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right).
\end{align}
We are therefore led to the coupled system of ODEs
\begin{align}
\label{eqn:Levo3}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo3}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k \, ,
\qquad k = -N_\iin/2, \ldots, N_\iin/2 -1
\end{align}
We solve system~\eqref{eq:mu}--\eqref{eqn:thetaevo3} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we choose the {\em midpoint} RK2 method so as to maximize the smoothing effect of the curvature term.

At the first stage of RK2 (the half step) we discretize \eqref{eq:mu}--\eqref{eqn:thetaevo3} as
\begin{align}
\label{eq:mu12d}
& \mu_k^{(n+1/2)} = \mu_k^{(n)} \exp \left( \eps k^2 \frac{\Dt}{2} \zeta^{(n)} \right) \\
\label{eq:L12d}
& \frac{L^{(n+1/2)} - L^{(n)}}{\Dt/2} = \Mterm^{(n)} \\
\label{eq:th12d}
& \frac{ \mu_k^{(n+1/2)} \thhat_k^{(n+1/2)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt/2} 
= \mu_k^{(n)} \hat{\NLterm}_k^{(n)}
\end{align}
where the superscript indicates the time step. At the next stage of RK2 it will be necessary to evaluate derivates at the half step $n+1/2$, thus requiring a call to the Stokes solver for which the body shapes must be known. We therefore insert \eqref{eq:mu12d} into \eqref{eq:th12d} and solve for $\theta^{(n+1/2)}$, giving
\begin{align}
\label{eq:th12}
\thhat_k^{(n+1/2)} = \left( \thhat_k^{(n)} + \frac{1}{2} \Dt \, \hat{\NLterm}_k^{(n)} \right)
\exp \left( -\epsilon k^2 \frac{\Dt}{2} \elfun^{(n)}\right) \, .
\end{align}
Since $\zeta >0$, the exponential term always acts to smooth the geometry and thus mitigate numerical instabilities. With the geometry at step $n+1/2$ known, a call to the Stokes solver allows the computation of $\Mterm^{(n+1/2)}$, $\widehat{\NLterm}_k^{(n+1/2)}$, and $\zeta^{(n+1/2)}$.

At the second stage of RK2, we discretize \eqref{eq:mu}--\eqref{eqn:thetaevo3} as 
\begin{align}
\label{eq:mu1d}
& \mu_k^{(n+1)} = \mu_k^{(n)} \exp \left( \eps k^2 {\Dt} \, \zeta^{(n+1/2)} \right) \\
\label{eq:L1d}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)}, \\
\label{eq:th1d}
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt} =
\mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}.
\end{align}
Once again, we must eventually compute the geometry at step $n+1$, and so we insert \eqref{eq:mu1d} into \eqref{eq:th1d} and solve for $\theta^{(n+1)}$, giving
\begin{align}
\thhat_k^{(n+1)} = & \thhat_k^{(n)} \exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) + \Dt \, \hat{\NLterm}_k^{(n+1/2)} \exp \left( - \frac{1}{2} \eps k^2 \Dt \left( 2 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right).
\end{align}
Again, both exponential terms act to smooth high-frequency oscillations in shape. We note that the above approach would generalize to higher-order Runge-Kutta methods, such as RK4.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Vorticity}
\label{sec:vorticity}
We start by computing the vorticity $w(\xx) = v_x - u_y$ for $\xx \in
\Omega$.  
\todo[inline]{Shouldn't we use $\omega$ (lower-case omega) for vorticity?}
As we did for the shear stress, we compute the vorticity due
to the double-layer potential, Stokeslets, and rotlets individually.
The vorticity of the double-layer potential~\eqref{eqn:velocityDLP} at
$\xx \in \Omega$ is
\begin{align}
  w^{\DD}(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} 
    \frac{(\rr \cdot \nn^\perp) + (\rr \cdot \nn)}{\rho^4}
    (\rr \cdot \eeta) ds_{\yy}.
  \label{eqn:vorticityDLP}
\end{align}
The vorticity due to the Stokeslet is
\begin{align*}
  w^S(\xx) = -\frac{1}{\pi} \sum_{\ell=1}^{M} 
    \frac{\rr \cdot \llambda_\ell^\perp}{\rho^2},
\end{align*}
and the vorticity of a rotlet is zero.  Therefore, the vorticity at $\xx
\in \Omega$ is
\begin{align*}
  w(\xx) = w^\DD(\xx) + w^S(\xx).
\end{align*}
To compute~\eqref{eqn:vorticityDLP} numerically, we apply the trapezoid
rule with an upsampling scheme described in Section~\ref{sec:NSI}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Drag}
\label{sec:drag}
With the pressure and shear stress, the drag on a single body with
boundary $\gamma_\ell$ is
\begin{align}
  F_D = \int_{\gamma_\ell} p \nn + \tau \ss \,ds.
  \label{eqn:drag}
\end{align}
The reader may notice an unusual sign convention for pressure in this formula, which is simply due to the normal direction pointing {\em into} the bodies.

@@

Figure \ref{01bodseq} shows the erosion of a single body, numerically
simulated with $N_\iin = 1024$ discretization points and shown at three
evenly spaced times. Here, color represents the vorticity of the
surrounding flow, which, when evaluated on the boundary, corresponds to the
shear stress and thus the local erosion rate. As seen in the left-most
snapshot, the body is initially circular and has the strongest vorticity
above and below it. The flow is for-aft symmetric as a result of the
time-reversibility of the Stokes equations, with vorticity vanishing at
the body's front and rear stagnation points. At the next instance, the
body has lost material due to erosion, but remains for-aft symmetric due
to the flow symmetry. The greatest material loss has occurred at the
body's top and bottom, where shear is strongest. Interestingly, the body
has developed well-defined corners at its front and rear stagnation
points. This geometric feature is a consequence of the erosion law,
which depends on the {\em absolute} shear stress: wherever the stress
vanishes, $\atau$ exhibits a corner, which manifests as a corner in the
geometry. As shown in the third snapshot, this corner persists as the
bodies becomes even more slender and continues to shrink. In this simulation,
the body completely vanishes at a final time of $t_f = 1.79E-2$.

To better understand the shape evolution, Fig.~\ref{shrink_intface}(a)
shows several superimposed interfaces at evenly spaced times. The
interfaces are color-coded by normalized time $t/t_f$. The body seems to
converge to a shape that resembles an American football --- a fairly
slender, for-aft symmetric shape with corners at the front and back.
Fig.~\ref{shrink_intface}(b) shows how the absolute shear stress,
$\atau$, is distributed along each of these interfaces. The normalized
arc length $s/L$ runs counterclockwise around the body, with $s/L = 0$
corresponding to the rear stagnation point and $s/L = 0.5$ the front
stagnation point. At early times (yellow to orange), the stress is
distributed non-uniformly, with highest values at the top and bottom of
the body. As time proceeds, though, the stress becomes more evenly
distributed along the body's surface (red). As the body nears the final
stages of erosion (dark red), the stress actually becomes slightly
higher near the stagnation points. A close look reveals that {\em at}
the stagnation points, the stress always vanishes, though in a region
that becomes increasingly narrow.

@@

A brief look at Fig.~\ref{shrink_intface}(a) reveals that
the spacing between successive interfaces becomes larger as the body
shrinks, suggesting that the overall shear stress is growing. Of course,
such a relationship is consistent with the well-known 3D Stokes drag
law, $\tau \sim \mu \umax / L$, since stress is inversely proportional
to body scale $L$. This law must be modified in two dimensions, though.
The modification is made non-trivial by subtleties of the 2D Stokes paradox, but can be correctly deduced by noting that, as the body vanishes, its drag must be consistent with slender-body theory. 
\todo[inline]{Here: add slender citations}
For a slender body in flow perpendicular to its long-axis $W \gg L$, the drag is proportional to
$\mu \umax W/ \log(W/L)$. This relationship implies that, for a 2D
cross-section, the stress per unit span can be estimated by
\begin{align}
  \label{eqn:stresslaw}
  \tau \sim \frac{\mu \umax}{L \log(W/L)}.
\end{align}
In interpreting our 2D simulations, we will treat the span, $W$, as fixed with the
requirement $W \gg L$.

The body's rate of area reduction is given by integrating the
absolute shear stress around the boundary
\begin{align*}
  \dot{A} = \int_{\gamma} \atau \, ds \, ,
\end{align*}
which gives the exact relationship
\begin{align*}
  \dot{A} = \mean{\atau} L \, .
\end{align*}
Thus, the area-loss rate is given by the product of the mean shear
$\mean{\atau}$ and the total perimeter $L$. Inserting the scaling
law~\eqref{eqn:stresslaw} gives
\begin{align*}
  \dot{A} \sim \frac{\mu \umax}{\log(W/L)} \, .
\end{align*}
Noting that $A \sim L^2$ allows one to solve the ODE to determine how the body area changes with time
\begin{align*}
  A (1 - \log A) = \areaconst (t_f - t)
\end{align*}
where $c_A$ is a constant.
This is an {\em implicit} relationship for how the area scales down in time during erosion.
In Fig.~6a, we show the body area as a function of time measured in the simulation
against the above scaling law with $\areaconst = 0.42$.
